# Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. The goal of this project is to use data from accelerometers on the belt, forearm, arm, and dumbbell of 6 participants as they perform barbell lifts correctly and incorrectly 5 different ways.
Six young healthy participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions:
* Class A - exactly according to the specification
* Class B - throwing the elbows to the front
* Class C - lifting the dumbbell only halfway
* Class D - lowering the dumbbell only halfway
* Class E - throwing the hips to the front
Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate. The exercises were performed by six male participants aged between 20-28 years, with little weight lifting experience. Researchers made sure that all participants could easily simulate the mistakes in a safe and controlled manner by using a relatively light dumbbell (1.25kg).

#Reference
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13). Stuttgart, Germany: ACM SIGCHI, 2013.

#Data 
The training data for this project are available at:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available at:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

#Goal
The goal of this project is to predict the manner in which subjects did the exercise. This is the "classe" variable in the training set. The model will use the other variables to predict with. This report describes:
* how the model is built
* use of cross validation
* an estimate of expected out of sample error

# Preliminary Work
## Reproduceability
An overall pseudo-random number generator seed was set at 1234 for all code. In order to reproduce the results below, the same seed should be used.
Different packages were downloaded and installed, such as caret and randomForest. These should also be installed in order to reproduce the results 

## Model Selection
Our outcome variable is classe, a factor variable with 5 levels. For this data set, "participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:
- exactly according to the specification (Class A)
- throwing the elbows to the front (Class B)
- lifting the dumbbell only halfway (Class C)
- lowering the dumbbell only halfway (Class D)
- throwing the hips to the front (Class E)
Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes.
Prediction evaluations will be based on maximizing the accuracy and minimizing the out-of-sample error. All other available variables after cleaning will be used for prediction.
Two models will be tested using decision tree and random forest algorithms. The model with the highest accuracy will be chosen as our final model.

## Cross-validation
Cross-validation will be performed by subsampling our training data set randomly without replacement into 2 subsamples: subTraining data (75% of the original Training data set) and subTesting data (25%). Our models will be fitted on the subTraining data set, and tested on the subTesting data. Once the most accurate model is choosen, it will be tested on the original Testing data set.

## Expected out-of-sample error
The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the subTesting data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

## Reasons for my choices
Our outcome variable "classe" is an unordered factor variable. Thus, we can choose our error type as 1-accuracy. We have a large sample size with N= 19622 in the Training data set. This allow us to divide our Training sample into subTraining and subTesting to allow cross-validation. Features with all missing values will be discarded as well as features that are irrelevant. All other features will be kept as relevant variables.
Decision tree and random forest algorithms are known for their ability of detecting the features that are important for classification. Feature selection is inherent, so it is not so necessary at the data preparation phase. Thus, there won't be any feature selection section in this report.

# Install packages and load the required libraries:

```{r}
library(lattice); 
library(ggplot2);
library(caret); 
library(randomForest); 
library(rpart); 
library(rpart.plot);
```

# Creating test and training data set.

```{r}
set.seed(1234)

# data load and clean up
trainingset <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!", ""))
testingset <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!", ""))

# Perform exploratory analysis - 
# dim(trainingset); dim(testingset); summary(trainingset); summary(testingset); str(trainingset); str(testingset); head(trainingset); head(testingset);               

# Delete columns with all missing values
trainingset<-trainingset[,colSums(is.na(trainingset)) == 0]
testingset <-testingset[,colSums(is.na(testingset)) == 0]

# Delete variables are irrelevant to our current project: user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7). 
trainingset   <-trainingset[,-c(1:7)]
testingset <-testingset[,-c(1:7)]


# partition the data so that 75% of the training dataset into training and the remaining 25% to testing
traintrainset <- createDataPartition(y=trainingset$classe, p=0.75, list=FALSE)
TrainTrainingSet <- trainingset[traintrainset, ] 
TestTrainingSet <- trainingset[-traintrainset, ]

# The variable "classe" contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the TrainTrainingSet data set and # compare one another.

plot(TrainTrainingSet$classe, col="red", main="Plot of levels of variable classe within the TrainTrainingSet data set", xlab="classe", ylab="Frequency")
```


Based on the graph above, we can see that each level frequency is within the same order of magnitude of each other. Level A is the most frequent while level D is the least frequent.

# Prediction model 1: Decision Tree
```{r}
model1 <- rpart(classe ~ ., data=TrainTrainingSet, method="class")
prediction1 <- predict(model1, TestTrainingSet, type = "class")
# Plot the Decision Tree
rpart.plot(model1, main="Classification Tree", extra=102, under=TRUE, faclen=0)
```

# Test results on our TestTrainingSet data set:
```{r}
confusionMatrix(prediction1, TestTrainingSet$classe)
```

# Prediction model 2: Random Forest
```{r}
model2 <- randomForest(classe ~. , data=TrainTrainingSet, method="class")

# Predicting:
prediction2 <- predict(model2, TestTrainingSet, type = "class")

# Test results on TestTrainingSet data set:
confusionMatrix(prediction2, TestTrainingSet$classe)
```

# Decision on which Prediction Model to Use:
Random Forest algorithm performed better than Decision Trees. Accuracy for Random Forest model was 0.995 compared to Decision Tree model with 0.739. The Random Forests model is choosen. The expected out-of-sample error is estimated at 0.005, or 0.5%.

# Submission
Here is the final outcome based on the Prediction Model 2 (Random Forest) applied against the Testing dataset
```{r}
# predict outcome levels on the original Testing data set using Random Forest algorithm
predictfinal <- predict(model2, testingset, type="class")
predictfinal
```
